---
title: "Introduction to aidR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to aidR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup, message = FALSE}
library(aidR)
library(dplyr)
```
## Overview

Data exported from automated insulin delivery (AID) systems are highly heterogeneous: each platform uses its own file format, variable names, data types, and units. The **aidR** package provides (i) functions to read exports of the most prevalent AID platforms, (ii) functions to standardize them into tidy, platform-independent formats for continuous glucose monitoring (CGM), basal insulin, bolus insulin, carbohydrate data, and self-monitored blood glucose (SMBG), and (iii) functions for visualization and conversion to other standardized formats. As a result, a downstream analysis can be written once and applied consistently across data from different AID systems.

The package supports the following steps:

- **Read** every export file belonging to one subject (`parse_data(clean = FALSE)`).
- **Clean** the most important data types (CGM, basal, bolus, carbohydrate, and SMBG) into an interchangeable format (`parse_data(clean = TRUE)`).
- **Filter** the merged data, e.g. to remove artefactual or duplicated CGM readings (`filter_CGM()`).
- **Check completeness** of the merged data, e.g. to flag missing data types or missing days (`check_completeness()`, `check_completeness_range()`), and verify that the insulin recorded in the individual data points matches the daily totals reported by the export (`check_insulin_totals()`).
- **Plot** daily data of one subject, with CGM, basal, bolus, carbohydrate, and SMBG data in one figure (`plot_day()`).
- **Merge across subjects** into one long-format data set for the whole study population (`merge_all()`).
- **Convert** the standardized data into other formats (e.g. [DIAX](https://github.com/Center-for-Diabetes-Technology/DIAX)) and into the layout expected by other R packages (e.g. [iglu](https://irinagain.github.io/iglu/) and [cgmquantify](https://CRAN.R-project.org/package=cgmquantify)).

The rest of this vignette walks through each step, describes the supported export formats, and shows how to filter, check, plot, and convert the resulting data.

## Installation

You can install aidR from CRAN with:
```{r, eval = FALSE}
install.packages("aidR")
```

## Supported export formats

aidR uses the *content* of each file to determine which platform it came from; it does not use the file name. You can therefore drop files or folders from several subjects and several platforms into one folder, and aidR will dispatch every file to the right reader automatically. Files with irrelevant extensions (e.g. `.png`, `.pdf`, `.jpg`, `.jpeg`, `.txt`, `.json`) are silently ignored.

aidR currently supports seven export formats, each described in more detail on its own page in this documentation:

| Platform                                           | Typical export                       | Supported AID systems (not exhaustive)                       |
|----------------------------------------------------|--------------------------------------|--------------------------------------------------------------|
| [Glooko](format-glooko.html)                       | folder with one CSV per data type    | CamAPS FX, Omnipod, ...                                      |
| [CareLink (Medtronic)](format-carelink.html)       | single `.xlsx` or `.csv` export      | MiniMed 780G, ...                                            |
| [mylife (Ypsomed)](format-mylife.html)             | single semicolon-delimited CSV       | CamAPS FX                                                    |
| [Tidepool](format-tidepool.html)                   | single multi-sheet `.xlsx` workbook  | Tandem Control-IQ, open-source DIY (e.g. Loop, AAPS), ...    |
| [YourLoops (Diabeloop)](format-yourloops.html)     | folder with one CSV per data type    | DBLG1, DBLG2                                                 |
| [Tandem Source](format-tandem-source.html)         | folder with one CSV per data type    | Tandem Control-IQ                                            |
| [Dexcom Clarity](format-dexcom-clarity.html)       | single delimited CSV (CGM only)      | Dexcom G6, G7, ... (any AID system using a Dexcom sensor)    |

## Reading and cleaning data for one subject

### Without cleaning

The main function is `parse_data()`, which expects a subject ID and the path(s) to the exported data (either relative, e.g. `./exports/my_export_file.csv`, or absolute, e.g. `~/Downloads/exports/my_export_file.csv`):
```{r, eval=FALSE}
data <- parse_data(id = "1", paths = "./exports/my_export_file.csv")
```
If several exports are available for one subject, e.g. for different time ranges, `paths` can be a vector holding all of the file names:
```{r, eval=FALSE}
data <- parse_data(id = "1", paths = c("./exports/my_export_file_id_1_month_1.csv", 
                                       "./exports/my_export_file_id_1_month_2.csv", 
                                       "./exports/my_export_file_id_1_month_3.csv"
                                       )
                   )
```
A path may point to a single file (e.g. a Tidepool or CareLink export), a folder (e.g. a Glooko or YourLoops export), or a zip archive. aidR detects this automatically, reads every file it finds within the path, and merges them into a single named list for that subject, with one entry per data type.

In the following example, we set `clean = FALSE`. This reads all of the files present in the export and applies minimal standardization (e.g. formatting timestamps), but performs no further cleaning or tidying. This is useful when you want to (i) inspect non-standard data types (e.g. alarms or pump settings rather than CGM), or (ii) simply read the data of a single subject, where merging across platforms does not matter.
```{r}
path <- system.file("extdata", "tidepool_example.xlsx", package = "aidR")
data_raw <- parse_data(id = "1", paths = path, clean = FALSE)
str(data_raw, max.level = 1)
```
Here, `path` happens to be a single Tidepool export, so `data_raw` contains the standard `cgm`, `basal`, `bolus`, and `SMBG` data frames alongside Tidepool-specific extras (`device_event`, `bolus_calculator`, `basal_schedules`, ...).

If a subject's data spans several files (i.e. `paths` is a vector of length > 1), `parse_data()` merges all of them together by data type and sorts each resulting data frame by `timestamp`. Exact duplicate rows within a data type are removed.

### With cleaning

Let us re-run the example above, but now with `clean = TRUE`. This is in fact the default, so it can be omitted.
```{r}
path <- system.file("extdata", "tidepool_example.xlsx", package = "aidR")
data_clean <- parse_data(id = "1", paths = path, clean = TRUE)
str(data_clean, max.level = 1)
```
The additional platform-specific data types from before (e.g. `device_event`, `bolus_calculator`) have disappeared. Instead, the major data types (CGM, basal, bolus, carbohydrates, and SMBG) are standardized to a common column layout.

All five start with the same four columns:

- `id`: the subject's id, as passed to `parse_data()`.
- `format`: the export format this row was parsed from, one of `glooko`, `carelink`, `mylife`, `tidepool`, `yourloops`, `tandem_source` or `dexcom_clarity`.
- `timestamp`: the time of the data point (`POSIXct`), in local (wall-clock) time.
- `timezone_offset`: the offset of that local time from UTC, in hours (`NA` if the export does not report it). UTC time is therefore `timestamp - lubridate::hours(timezone_offset)`.

Rows are sorted by `timestamp`, and the columns that follow are specific to the data type:

- **`cgm`**: CGM values, one row per sensor reading
    - `value`: the measured CGM value, always in mg/dL (readings exported in mmol/L are converted).
    - `unit`: the unit of `value` (always `mg/dL`).
    - `sensor_name`: the name of the CGM sensor, `NA` if not available.
- **`basal`**: delivered basal rates, one row per interval over which the rate was constant
    - `duration`: the duration of the basal rate, in hours, starting at `timestamp`.
    - `rate`: the delivered basal rate, in U/h. The insulin delivered over the interval is `rate * duration`.
    - `unit`: the unit of `rate` (always `U/h`).
    - `pump_name`: the name of the pump, `NA` if not available.
- **`bolus`**: bolus information, one row per bolus
    - `type`: the type of the bolus. Either `normal` (delivered in one go), `square_wave` (delivered gradually and evenly over a longer period), `dual_wave` (part delivered immediately, the rest gradually over a longer period), `biphasic` (YourLoops only, a bolus split into two boluses, separated by a period of time; each part is its own row), or `pen` (YourLoops only).
    - `total`: the total amount of insulin in the bolus (`normal` + `extended`), in U.
    - `normal`: the portion delivered immediately, in U (for `normal`, `dual_wave` and `biphasic`; `0` for `square_wave`).
    - `extended`: the portion delivered gradually over a longer period, in U (for `dual_wave` and `square_wave`, `0` otherwise).
    - `unit`: the unit of `total`, `normal` and `extended` (always `U`).
    - `duration_extended`: the duration of the extended portion, in hours (for `dual_wave` and `square_wave`, `NA` otherwise, and also `NA` where the platform does not report it).
    - `pump_name`: the name of the pump, `NA` if not available.
- **`carbs`**: user-reported carbohydrate consumption, one row per entry
    - `value`: the carbohydrate value (in g).
    - `unit`: the unit of `value` (always `g`).
    - `label`: an optional label for the entry, `NA` if not available. Its content is platform-specific: it may be a meal name entered by the user, or summarize additional information on the entry (e.g. that it was a fatty meal).
    - `estimated_absorption_duration`: the absorption duration in hours, as estimated by the user, `NA` if not available.
    - `is_hypo_treatment`: a logical flagging entries labeled as hypoglycemia treatment, `NA` if not available.
- **`SMBG`**: self-monitored blood glucose, one row per measurement
    - `value`: the measured SMBG value, always in mg/dL (readings exported in mmol/L are converted).
    - `unit`: the unit of `value` (always `mg/dL`).
    - `sensor_name`: the name of the SMBG device, `NA` if not available.

This layout is the same for all platforms and can therefore be merged across subjects. Note that a column being present does not mean every platform fills it: for example, `timezone_offset`, `sensor_name` and `pump_name` in particular are only populated where the export reports them.

In addition, two data types hold the daily insulin totals, with one row per day rather than one row per data point:

- **`total_basal`**: `id`, `format`, `date`, `total_basal`, and `source`
- **`total_bolus`**: `id`, `format`, `date`, `total_bolus`, and `source`

`total_basal` and `total_bolus` give the insulin delivered on that day, in units. Where a platform reports a daily aggregate itself, that value is used and `source` is `"reported"`; otherwise the total is summed over the cleaned `basal` (as `rate * duration`) or `bolus` data and `source` is `"summed"`:

| Format | `total_basal` | `total_bolus` |
|---------------|------------|------------|
| Glooko        | reported   | reported   |
| Tandem Source | reported   | summed     |
| CareLink      | -          | summed     |
| Tidepool      | summed     | summed     |
| YourLoops     | summed     | summed     |
| mylife        | -          | summed     |
| Dexcom Clarity | -         | -          |

This distinction matters because a summed total can never be more complete than the underlying records. For Omnipod users on Glooko, for example, the basal field contains only the basal rates delivered in manual mode, because rates delivered in automated mode are not exported. Summing this field would therefore substantially understate the insulin actually delivered. The reported daily aggregate, however, does cover automated mode. This aggregate can therefore be used later as a check on the completeness of the export.

Note that there is no `total_basal` for mylife, since a mylife export contains no basal data at all, and none for CareLink, since a CareLink export reports only the *scheduled* basal rates rather than the basal insulin effectively delivered. For the same reason, neither platform yields a cleaned `basal` data type. A Dexcom Clarity export carries neither, since it is a CGM export and contains no insulin data at all.

## Filtering

AID data is often noisy and may require additional filtering. While we encourage users to inspect the parsed data for platform-specific issues, aidR implements a few filters that we have found particularly useful. CGM data in particular is frequently affected in two ways: individual values may be physiologically implausible (below or above the range a sensor can measure), and the same period may be covered by two overlapping data streams whose timestamps are shifted by a few seconds or minutes but that otherwise carry the same signal (often the case for DIY open-source systems).

`filter_CGM()` addresses both. It first truncates values to a plausible range, clamping anything below `truncate_min_cgm` (40 mg/dL by default) up to that value and anything above `truncate_max_cgm` (400 mg/dL by default) down to it. It then downsamples the CGM series to a regular time grid (5-minute bins by default): each reading is assigned to a time bin, and within each bin only the reading closest to the bin's start is kept. Two streams that differ only by a shift of a few seconds or minutes therefore fall into the same bins and collapse to a single value, so a duplicated stream is removed rather than doubling the number of readings for that period.
```{r}
nrow(data_clean$cgm)
data_clean_filter <- filter_CGM(data_clean)
nrow(data_clean_filter$cgm)
```
The two truncation bounds and the downsampling interval can each be adjusted or disabled: `truncate_min_cgm` (default 40 mg/dL), `truncate_max_cgm` (default 400 mg/dL), and `min_minutes_difference_cgm` (default 5 minutes); set any of them to `NULL` to skip that particular step. At present only CGM is filtered, and only once it has been cleaned (i.e. after `parse_data(clean = TRUE)`); all other data types are returned unchanged.

## Checking data completeness

Before running an analysis, it is often useful to verify that a subject actually has data of every type you need, and that it is not missing entire days. `check_completeness()` warns if a data type is absent, or present but empty:
```{r}
check_completeness(data_clean)
```
`check_completeness_range()` goes a step further and checks, for a given date range, which specific days have no data at all for each type. This is useful for spotting gaps such as a sensor or pump disconnecting for a few days, or users that exported data from a wrong time range:
```{r}
check_completeness_range(
  data_clean,
  first_date = "2019-01-10",
  last_date  = "2019-05-31"
)
```
In this case, it appears that there is a gap in all data types spanning from 2019-04-15 until 2019-04-19, and additionally no CGM data for 2019-05-21.

Both functions accept a `necessary_types` argument to restrict (or extend) which data types are checked. They also accept the data of several subjects at once (the output of `merge_all()`, see below), in which case every subject is checked separately and the warnings name the id they refer to.

## Checking the daily insulin totals

`check_insulin_totals()` asks whether the reported insulin adds up: it compares the insulin summed over the individual `basal` entries against the daily totals in `total_basal`, and warns for every day where the two disagree. Per default, only the basal insulin is checked - the argument `types` extends the check to the bolus insulin (`types = c("basal", "bolus")`, comparing `bolus` against `total_bolus`).

This is only informative where the daily total was reported by the export itself, i.e. where `source == "reported"`. Where `aidR` summed the total itself (`source == "summed"`), the two agree by construction and are skipped. In practice, this means the check is useful for Glooko (basal and bolus) and for Tandem Source (basal only).

A mismatch means the individual data points do not cover the whole day. The Glooko example export shows the typical picture for an Omnipod user:
```{r}
glooko_path <- system.file("extdata", "glooko_example.zip", package = "aidR")
data_glooko <- parse_data(id = "1", paths = glooko_path)
check_insulin_totals(data_glooko)
```
The warning names the days that disagree. As with the completeness checks, the data of several subjects can be passed at once, and each is then checked separately. To see *how far* they disagree, use the comparison that the function returns invisibly - here for both insulin types:
```{r}
totals <- suppressWarnings(
  check_insulin_totals(data_glooko, types = c("basal", "bolus"))
)
totals
```
The bolus totals agree exactly, while the summed basal is a small fraction of the reported one (see column `within_tolerance` which types and dates fail the comparison). This is the signature of an Omnipod user: the export records the basal rates delivered in manual mode only, so almost all of the basal insulin - everything delivered in automated mode - is missing from `basal`, but is still included in `total_basal`. For such subjects, an analysis based on `basal` would largely understate the insulin actually delivered, whereas `total_basal` remains correct.

## Plotting single days

`plot_day()` plots one day of a subject's data as two stacked panels: the CGM trace over the shaded target range with the SMBG measurements as diamonds, and the insulin and carbohydrate information below it - the basal rate as a filled step area, the boluses as labelled circles, and the carbohydrates as labelled boxes, with the insulin delivered over the whole day given on top:
```{r, fig.width=9, fig.height=6.5, out.width="100%"}
plot_day(data_clean, id = "1", days = "2019-04-01")
```
The argument `days` may take several days at once. Per default, every day that has CGM data is plotted (one plot per day).

Glucose units can be specified with the argument `units` as `"mg/dl"` (default) or `"mmol/l"`.

A data type (except CGM) that is absent is simply left out, so the function also works for exports that carry no basal data at all, such as CareLink. If many small boluses were delivered, this can be hard to read - in this case, the argument `min_bolus` can be used to de-clutter the plot: boluses below `min_bolus` are then only marked by a small tick at their time, without a circle and without their value, and a note next to the daily totals says so.

## Merging data across subjects

Once every subject's data has been parsed and cleaned, `merge_all()` combines them into one long-format data set per data type:
```{r}
path_2 <- system.file("extdata", "glooko_example.zip", package = "aidR")
data_2  <- parse_data(id = "2", paths = path_2) # Note: clean = TRUE by default

all_data <- list("1" = data_clean, "2" = data_2)

combined <- merge_all(all_data)

str(combined, max.level = 1)
head(combined$cgm)
tail(combined$cgm)
```
By default, `merge_all()` returns the five core types (`cgm`, `basal`, `bolus`, `carbs`, `SMBG`); a subject missing one of them simply contributes no rows to that type. The `types` argument restricts which data types are merged.

## Converting to other formats

Three helper functions convert the standardized data into the formats expected by other tools.

`get_iglu_format()` reshapes the `cgm` data frame into the `id`/`time`/`gl` layout expected by the [iglu](https://irinagain.github.io/iglu/) package, so that CGM metrics and visualizations from iglu can be used directly:
```{r, comment="", fig.width=12, fig.height=7, out.width="100%", warning=FALSE}
iglu_1 <- get_iglu_format(data_clean)
head(iglu_1)
iglu::plot_agp(iglu_1)
```

`get_cgmquantify_format()` builds the data frame that the metric functions of the [cgmquantify](https://CRAN.R-project.org/package=cgmquantify) package expect, with columns `Time`, `glucose`, `Date`, `time_of_day`, and `type_of_event`. This is the same layout that `cgmquantify::readfile()` produces from a Dexcom export, so the result can be handed to any cgmquantify function directly and no intermediate file has to be written:
```{r, comment="", eval = requireNamespace("cgmquantify", quietly = TRUE)}
cgmq_1 <- get_cgmquantify_format(data_clean)
head(cgmq_1)
cgmquantify::interdaycv(cgmq_1)
cgmquantify::J_index(cgmq_1)
```

Note that cgmquantify applies to a single individual only, and `get_cgmquantify_format()` requires the `id` argument as soon as `data` holds more than one `id`.

`write_DIAX_format()` writes CGM, basal, bolus, carbs, and SMBG data to one JSON file per subject, following the [DIAX](https://github.com/Center-for-Diabetes-Technology/DIAX) specification. It accepts either the output of `parse_data()` for a single subject or the output of `merge_all()` for several, and writes one file per id to `paste0(prefix, id, ".json")`:
```{r}
diax_dir <- tempfile()
dir.create(diax_dir)
write_DIAX_format(combined, prefix = file.path(diax_dir, "diax_id_"))
list.files(diax_dir)
```

## Putting it all together

A typical study-level pipeline combines all of the above for every subject in a loop (note the `options(warn = 1)` to show all warnings immediately rather than at the end of the loop):
```{r, eval = FALSE}
# Define a data frame: first column is id, second column the export paths
# An id may appear multiple times if multiple exports (e.g. across many months) are available
# Note: filenames are not used to determine the file type and/or id
input <- data.frame(id = c("1", "1", "1", "2", "3", "3", "4", "5"),
                    paths = c(
                      "./tidepool_id_1_period_1.xlsx",
                      "./tidepool_id_1_period_2.xlsx",
                      "./tidepool_id_1_period_3.xlsx",
                      "./glooko_period_1.zip",
                      "./yourloops_1.zip",
                      "./yourloops_2.zip",
                      "./tidepool_id_4_period_1.xlsx",
                      "./carelink.csv"
                      )
                    )
ids <- unique(input$id)
all_data <- list()

options(warn = 1) # warn immediately

for (id in ids) {
  all_data[[id]]          <- parse_data(id, input$paths[input$id == id])
  all_data[[id]]          <- filter_CGM(all_data[[id]])
  check_completeness(all_data[[id]])
  check_insulin_totals(all_data[[id]])
}
combined <- merge_all(all_data)
# Inspect a single day of a single subject
plot_day(combined, id = "1", days = "2026-06-22")
# write DIAX files
diax_dir <- "./diax"
write_DIAX_format(combined, prefix = file.path(diax_dir, "diax_id_"))
# Plot using iglu
iglu_combined <- get_iglu_format(combined)
iglu::plot_glu(iglu_combined, subjects = "2")
```
`combined` can then be filtered, summarized, converted to other formats, or exported for statistical analysis exactly as shown in the previous section.