---
title: "aidR: mylife export format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{aidR: mylife export format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(aidR)
```

## Overview

A mylife (Ypsomed) export is a *single* semicolon-delimited `.csv` file with
eight columns, holding one row per logged event. Unlike the platforms that ship
one file per data type, all event types are interleaved in the same file and are
distinguished only by the value of the event column, e.g.

```
"Tag";"Datum";"Uhrzeit";"Ereignis";"Wert";"Einheit";"Information";"Notiz"
"Fr.";"16.01.26";"09:05";"Kohlenhydrate";"42";"g KH";;
"Fr.";"16.01.26";"09:04";"Bolus";"4,40";"U";"Bolusvorschlag: 4.40 U";
```

Both the column headers and the event labels are language-specific and differ
with the user's account language (the example above is German). Numeric values
use a comma as the decimal separator (`4,40` = 4.40), the date and time are held
in separate columns, and additional context is only available as free text in
the information column.

```{r}
path <- system.file("extdata", "mylife_example.csv", package = "aidR")
```

## How aidR detects this format

Because both headers and event labels are language-specific, `aidR` cannot rely
on their text to identify a mylife file. Instead it checks the *structure* of
the header row, namely that it splits into exactly eight fields. For the same
reason, event types are not matched by label either: bolus rows are recognized
by their unit starting with `U`, and carbohydrate rows by their unit starting
with `g`, both of which are stable across languages.

## Entries and the data types they produce

```{r}
data <- parse_data(id = "1", paths = path)
```

`aidR` compiles the raw mylife entries into a smaller set of data types. The
list below describes each entry type, shows an example and, in italics, notes
which clean data type(s) it is compiled into.

- **Bolus entries** (unit `U`; labelled e.g. `Bolus`): insulin boluses.
  *Compiled into `bolus`.*
```{r}
head(data$bolus)
```
- **Carbohydrate entries** (unit `g`; labelled e.g. `Kohlenhydrate`):
  carbohydrate intake communicated to the pump. *Compiled into `carbs`.*
```{r}
head(data$carbs)
```
- **All other entry types**: preserved as-is under their original,
  language-specific label, so nothing in the export is lost even though it is
  not standardized. In the bundled German example these are the raw `Bolus` and
  `Kohlenhydrate` groups the two clean types above were built from:
```{r}
setdiff(names(data), c("bolus", "carbs"))
```

Note that a mylife export contains **no CGM, basal, or SMBG data** - `bolus` and
`carbs` are the only core data types available from this platform.

## Notes

- `estimated_absorption_duration` is parsed out of the free-text information
  column (e.g. `"... 2h ..."`) and is `NA` when not stated.
- `is_hypo_treatment` is likewise derived from the information column, by
  looking for the marker `HypoTreatment`.
- `label` retains the original, language-specific event label, so the source of
  each carbohydrate entry stays traceable after merging across platforms.
- `timezone_offset` is always `NA`, since mylife exports report only local
  (wall-clock) time.
