---
title: "aidR: CareLink export format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{aidR: CareLink export format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(aidR)
```

## Overview

A CareLink (Medtronic) export is a *single* semicolon-delimited `.csv` (or
`.xlsx`) file that bundles everything into one sheet. Rather than one file per
data type, the file is divided into **sections** separated by marker rows of the
form `-------;MiniMed 780G MMT-1111;Pump;XXXXXXXXXX;-------`. Each section has
its own header row and its own set of columns, and within a section each row
holds only the few columns relevant to its event type - so a single section
mixes boluses, basal rates, alerts, and more, with most columns empty on any
given row.

Numeric values use a comma as the decimal separator (`1,9` = 1.9), and
timestamps are split across separate `Date` and `Time` columns.

```{r}
path <- system.file("extdata", "carelink_example.csv", package = "aidR")
```

## How aidR detects this format

`aidR` looks for the characteristic multi-section layout, i.e. the presence of
`-------` marker rows. Because everything arrives in one file, most of the
parsing work happens up front: the file is split into sections, and the large
"Pump" section is then split further into individual event types based on which
columns are populated on each row. As a safety net, `aidR` checks that no
column containing data was missed while splitting, so a change in Medtronic's
export layout surfaces as an error rather than silently dropped data.

## Sections and the data types they produce

```{r}
data <- parse_data(id = "1", paths = path)
```

`aidR` compiles the raw CareLink sections into a smaller set of data types. The
list below describes each section, shows an example and, in italics, notes
which clean data type(s) it is compiled into.

- **Header block** (first rows, before the first `-------`): patient, pump, and
  sensor information. Not a data type of its own; `aidR` reads the pump name
  (e.g. `MiniMed 780G MMT-1111`) and sensor name (e.g. `Guardian™ 4 Sensor`)
  from here and attaches them to the relevant rows as `pump_name` and
  `sensor_name`.
- **Pump section**: the bulk of the export. Split by event type into:
  - basal rates. *Compiled into `basal`.*
```{r}
head(data$basal)
```
  - boluses. *Compiled into `bolus`.* Records whether a bolus was normal,
    square-wave, or dual-wave, along with the delivered volume and, for extended
    boluses, the duration.
```{r}
head(data$bolus)
```
  - bolus wizard (BWZ) entries, i.e. the carbohydrates communicated to the bolus
    calculator together with the resulting recommendation. *Compiled into
    `carbs`.* `aidR` matches these against the bolus events so that a
    carbohydrate entry can be linked to the bolus it triggered.
```{r}
head(data$carbs)
```
  - alerts and alarms raised by the pump.
```{r}
head(data$alerts)
```
  - prime events (tubing/cannula fills).
```{r}
head(data$prime)
```
  - additionally, where present: temp basal events, correction-bolus
    information, sensor calibrations, insulin action curve changes, and device
    change events. These are preserved under their own names but are absent from
    the bundled example, since the sections are empty for this participant.
- **Aggregated Auto Insulin Data section**: daily aggregated insulin delivered
  in automated mode.
```{r}
head(data$aggr_daily_insulin)
```
- **Sensor section**: CGM data. *Compiled into `cgm`.* Sensor exceptions
  (e.g. calibration required, sensor warm-up) are extracted separately into
  `sensor_exceptions` where present.
```{r}
head(data$cgm)
```

Self-measured blood glucose (SMBG) is also read from the pump section and
compiled into `SMBG` when the export contains meter readings; the bundled
example has none, so that entry is absent here.

## Notes

- CGM and SMBG values are reported in mg/dL or mmol/L depending on account
  settings; `aidR` converts to mg/dL automatically.
- `timezone_offset` is always `NA`, since CareLink exports report only local
  (wall-clock) time.
- Which sections are present depends on the pump/CGM combination and on what the
  participant actually used; empty sections are dropped rather than returned as
  empty data frames.
