---
title: "aidR: mylife export format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{aidR: mylife export format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup, message = FALSE}
library(aidR)
```
```{r, include=FALSE}
path <- system.file("extdata", "mylife_example.csv", package = "aidR")
```
## Overview

A mylife export is a *single* semicolon-delimited `.csv` file with eight columns:
```{r, comment = "", echo = FALSE}
writeLines(readLines(path, n = 10))
```
Each event represents a single row. Both the column headers and the event labels are language-specific and vary with the language of the user's account (the example above is in German).

## How aidR detects this format

Because the headers in a mylife export are language-specific, `aidR` cannot identify a file from its header names. It relies instead on the *structure* of the header row, which is delimited by semicolons `;` and must split into exactly eight columns. For the same reason, event types are not matched by their labels either: bolus rows are recognized by a unit starting with `U`, and carbohydrate rows by a unit starting with `g`.

## Entries and the data types they produce

To inspect the raw entries, we parse the export with `clean = FALSE`, which returns the full information provided by mylife:
```{r}
data <- parse_data(id = "1", paths = path, clean = FALSE)
```
All entries are preserved as-is under their original, language-specific label, so nothing in the export is lost. Possible labels include:

- **Bolus entries**: insulin boluses. German label e.g. `Bolus`; unit `U`.
```{r, comment=""}
head(data$Bolus)
```
- **Carbohydrate entries**: carbohydrate intake communicated to the pump. German label e.g. `Kohlenhydrate`; unit `g`.
```{r, comment=""}
head(data$Kohlenhydrate)
```
- **Manual blood glucose entries**: self-monitored blood glucose (SMBG). German label e.g. `Blutzucker manuelle Eingabe`; unit `mg/dL` or `mmol/L`.
```{r, comment=""}
head(data$`Blutzucker manuelle Eingabe`)
```
- **Ease-off** (`Ease-off`): start and duration of Ease-off mode.
- **Boost** (`Boost`): start and duration of Boost mode.
- **Hypo carbohydrate entries**: carbohydrates taken to treat hypoglycemia. German label e.g. `Hypo Kohlenhydrate`; unit `g`.
- **Auto mode on**: start and duration of Auto mode. German label e.g. `Auto mode An`.
- **Slowly resorbing**: carbohydrate intake recorded as slowly resorbing. German label e.g. `langsam resorbierbar`; unit `g`.
- **Auto mode loading**: start and end of Auto mode loading, including the reason for any interruption. German label e.g. `Auto mode lädt`.

Note that a mylife export contains **no CGM or basal data** - `bolus`, `carbs`, and `SMBG` are the only core data types available from this platform.

## Cleaning files

When `parse_data` is run with `clean = TRUE`, the cleaned data types are built from the following raw files and attributes:
 
| Cleaned data type | Used files           | Used attributes                                                                                                                |
|-------------------|----------------------|--------------------------------------------------------------------------------------------------------------------------------|
| `bolus`           | `mylife_example.csv` | All entries whose unit is `U`, i.e. bolus entries.                                                                             |
| `carbs`           | `mylife_example.csv` | All entries whose unit is `g`, including regular carbohydrates, slowly resorbing carbohydrates, and hypo carbohydrate entries. |
| `SMBG`            | `mylife_example.csv` | All entries whose unit is `mg/dL` or `mmol/L`, i.e. manual blood glucose entries.                                              |
| `total_bolus`     | `mylife_example.csv` | The cleaned `bolus` data, summed per day.                                                                                      |

## Notes

- SMBG values may be reported in mg/dL or mmol/L depending on the account settings; `aidR` converts them to mg/dL automatically.
- `timezone_offset` is always `NA`, since mylife exports report only local (wall-clock) time.
- There is no `total_basal`, since a mylife export contains no basal data. `total_bolus` is summed from the cleaned `bolus` data and carries `source = "summed"`.
- `estimated_absorption_duration` is parsed out of the free-text information column for slowly resorbing carbs (e.g. `"... 2h ..."`) and is `NA` when not stated.
- `is_hypo_treatment` is likewise derived from the information column, by looking for the marker `HypoTreatment`.
- `label` retains the original, language-specific event label, so the source of each carbohydrate entry stays traceable after merging across platforms.