---
title: "aidR: Tidepool export format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{aidR: Tidepool export format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(aidR)
```

## Overview

A Tidepool export is a *single* multi-sheet `.xlsx` workbook with one sheet per
data type. The sheets `CGM`, `Basal`, and `Bolus` are always expected to be
present; the remaining sheets (`Device Event`, `Food`, `Bolus Calculator`,
`Basal Schedules`, `BG Targets`, `Carb Ratios`, `Insulin Sensitivities`, `SMBG`,
`Upload`) are optional and are read only if the workbook contains them.

Because Tidepool aggregates data uploaded from many different pumps, meters, and
CGM devices, the exact set of sheets and the columns within them vary between
exports. Unlike Glooko or mylife, sheet names and headers are always in English.

```{r}
path <- system.file("extdata", "tidepool_example.xlsx", package = "aidR")
```

## How aidR detects this format

`aidR` checks that the file is a readable Excel workbook containing a sheet
named `CGM`. Since Tidepool's sheet names are stable and language-independent,
no content-based heuristics are needed - each sheet is looked up by name, and
missing optional sheets simply yield no data rather than an error.

## Sheets and the data types they produce

```{r}
data <- parse_data(id = "1", paths = path)
```

`aidR` compiles the raw Tidepool sheets into a smaller set of data types. The
list below describes each sheet, shows an example and, in italics, notes which
clean data type(s) it is compiled into.

- **`CGM`**: CGM data. *Compiled into `cgm`.*
```{r}
head(data$cgm)
```
- **`Basal`**: basal rates, including the delivery type (e.g. scheduled versus
  temporary). *Compiled into `basal`.*
```{r}
head(data$basal)
```
- **`Bolus`**: insulin boluses. *Compiled into `bolus`.*
```{r}
head(data$bolus)
```
- **`SMBG`**: self-measured blood glucose. *Compiled into `SMBG`.* The raw sheet
  is additionally preserved as `smbg`.
```{r}
head(data$SMBG)
```
- **`Bolus Calculator`** or **`Food`**: carbohydrate intake. *Compiled into
  `carbs`.* Tidepool exports carry one or the other depending on the uploading
  device, never both, and `aidR` uses whichever is present. If both or neither
  appear, `aidR` raises an error rather than guessing - so a change in
  Tidepool's export format is caught immediately. The bundled example contains a
  `Bolus Calculator` sheet, which is also preserved raw as `bolus_calculator`.
```{r}
head(data$carbs)
```
- **`Device Event`**: device-level events such as alarms, suspends, resumes, and
  cannula/reservoir changes.
```{r}
head(data$device_event)
```
- **`Basal Schedules`**, **`BG Targets`**, **`Carb Ratios`**,
  **`Insulin Sensitivities`**: pump therapy settings as they changed over time.
  These are preserved under their own names, since their structure is
  pump-specific and has no equivalent on other platforms.
```{r}
head(data$basal_schedules)
```
- **`Upload`**: metadata about each upload to Tidepool (device, time, software
  version), useful for tracing which device a given stretch of data came from.
```{r}
head(data$upload)
```

## Notes

- The `CGM` sheet is read with a generous `guess_max`, because columns that are
  sparse early in a large sheet would otherwise be assigned the wrong type.
- Timestamps carry a per-row UTC offset, so unlike most other platforms
  `timezone_offset` is populated rather than `NA`.
- CGM and SMBG values are converted to mg/dL automatically where reported in
  mmol/L.
- Some workbooks contain an `EXPORT ERROR` sheet listing records Tidepool could
  not export; `aidR` does not read it, so consult it directly if data appear to
  be missing.
