---
title: "aidR: CareLink export format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{aidR: CareLink export format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(aidR)
```

## Overview

```{r, include=FALSE}
path <- system.file("extdata", "carelink_example.csv", package = "aidR")
```

A CareLink (Medtronic) export is a *single* semicolon-delimited `.csv` (or `.xlsx`) file. The file is divided into *sections* separated by marker rows of the form `-------;MiniMed 780G MMT-1886;Pump;XXXXXXXXXX;-------`. Each section has its own header row and its own set of columns:

```{r, comment = "", echo = FALSE}
writeLines(readLines(path, n = 10))
```

## How aidR detects this format

`aidR` identifies a CareLink export by the keyword `MiniMed`: CareLink only accepts uploads from Medtronic MiniMed pumps, and MiniMed pumps can only upload to CareLink, so the presence of `MiniMed` uniquely marks the format.

## Sections and the data types they produce

To inspect the files, we parse the export with `clean_files = FALSE`, which returns the full information provided by CareLink:

```{r}
data <- parse_data(id = "1", paths = path, clean_files = FALSE)
```

Because all information is present in one file, most of the parsing work happens up front: the file is split into sections, and the large "Pump" section is then split further into individual event types based on which columns are populated on each row. 

The list below describes the differnt sections and the information we can get from it.

- **Header block** (first rows, before the first `-------`): patient, pump, and sensor information. `aidR` reads the pump name (e.g. `MiniMed 780G MMT-1886`) and sensor name (e.g. `Guardian™ 4 Sensor`) from here.

- **Pump section**: split by event type into

  - basal rates.

```{r}
head(data$basal)
```

  - boluses. Records whether a bolus was normal, square-wave, or dual-wave, along with the delivered volume and, for extended boluses, the duration.

```{r}
head(data$bolus)
```

  - bolus wizard (BWZ) entries, i.e. the carbohydrates communicated to the bolus calculator together with the resulting recommendation. `aidR` matches these against the bolus events so that a
    carbohydrate entry can be linked to the bolus it triggered.

```{r}
head(data$bwz)
```

  - self-measured blood glucose (SMBG) entries.

```{r}
head(data$SMBG)
```

  - alerts and alarms raised by the pump.

```{r}
head(data$alerts)
```

  - prime events (tubing/cannula fills).

```{r}
head(data$prime)
```

  - temp basal events.

```{r}
head(data$temp_basal)
```

  - correction-bolus information (not present in this example).

```{r}
head(data$corr_bolus_info)
```

  - sensor calibrations.

```{r}
head(data$sensor_calibration)
```

  - insulin action curve changes.

```{r}
head(data$insulin_action_curve)
```

  - device change events.

```{r}
head(data$device_changes)
```

- **Aggregated Auto Insulin Data section**: daily aggregated insulin delivered in automated mode.

```{r}
head(data$aggr_daily_insulin)
```

- **Sensor section**: CGM data.

```{r}
head(data$cgm)
```

## Cleaning files

When `parse_data` is run with `clean_files = TRUE`, the cleaned data types are built from the following raw files and attributes:

| Cleaned data type |  Used files            | Used attributes |
|-------------------|------------------------|-----------------|
| `cgm`             | `carelink_example.csv` | `data$cgm`      |
| `basal`           | `carelink_example.csv` | `data$basal`    |
| `bolus`           | `carelink_example.csv` | `data$bolus`    |
| `carbs`           | `carelink_example.csv` | `data$bwz`      |
| `SMBG`            | `carelink_example.csv` | `data$SMBG`     |

## Notes

- CGM and SMBG values are reported in mg/dL or mmol/L depending on account settings; `aidR` converts to mg/dL automatically.
- `timezone_offset` is always `NA`, since CareLink exports report only local (wall-clock) time.

