---
title: "aidR: YourLoops export format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{aidR: YourLoops export format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(aidR)
```

## Overview

A YourLoops (Diabeloop) export is a folder (possibly zipped) containing one CSV
per data type, named `cbg.csv`, `basal.csv`, `bolus.csv`, `meals.csv`,
`rescuecarbs.csv`, `alarms.csv`, `physicalActivities.csv`, plus several
`settings-*.csv` files describing the CGM, pump, and device configuration.

Unlike Glooko or mylife, the column headers are always in English regardless of
the user's account language. Timestamps are given in UTC alongside a separate
timezone offset in minutes, so recovering local (wall-clock) time requires
combining the two.

```{r}
path <- system.file("extdata", "yourloops_example.zip", package = "aidR")
```

## How aidR detects this format

Because YourLoops headers are stable and language-independent, `aidR` identifies
each file by matching its column names exactly against a lookup of the expected
headers per data type. The column-name sets are unique across data types, so a
file matches at most one type; the filename itself is never used.

## Files and the data types they produce

```{r}
data <- parse_data(id = "1", paths = path)
```

`aidR` compiles the raw YourLoops CSVs into a smaller set of data types. The
list below describes each raw file, shows an example and, in italics, notes
which clean data type(s) it is compiled into.

- **`cbg.csv`**: CGM data ("continuous blood glucose"), including the CGM model.
  *Compiled into `cgm`.*
```{r}
head(data$cgm)
```
- **`basal.csv`**: basal rates, with the duration given in milliseconds and the
  delivery type. *Compiled into `basal`.*
```{r}
head(data$basal)
```
- **`bolus.csv`**: insulin boluses, including the amount required versus
  delivered and the insulin on board. *Compiled into `bolus`.*
```{r}
head(data$bolus)
```
- **`meals.csv`**: carbohydrate intake communicated to the bolus calculator.
  *Compiled into `carbs`.*
- **`rescuecarbs.csv`**: carbohydrates taken to treat a hypoglycaemic event.
  *Also compiled into `carbs`*, flagged via `is_hypo_treatment = TRUE`, so both
  sources end up in one data frame while remaining distinguishable.
```{r}
head(data$carbs)
```
- **`alarms.csv`**: alarms raised by the loop or the pump, with severity level
  and acknowledgement status.
```{r}
head(data$alarms)
```
- **`physicalActivities.csv`**: self-reported physical activity, with duration
  and reported intensity. Often sparse or missing entirely.
```{r}
head(data$physicalActivities)
```
- **`settings-cgm.csv`**, **`settings-pump.csv`**, **`settings-device.csv`**,
  **`settings-current-parameters.csv`**, **`settings-security-basals.csv`**:
  the CGM, pump, smartphone, loop-parameter, and security-basal configuration.
  These are preserved under their own names (e.g. `settings_pump`), since they
  describe the system setup rather than time series data.
```{r}
head(data$settings_pump)
```

Note that a YourLoops export contains **no SMBG data**.

## Notes

- Timestamps are reported in UTC together with a `timezoneOffSet` in minutes.
  `aidR` adds the offset to obtain local (wall-clock) time in `timestamp`, and
  reports the offset itself in hours as `timezone_offset`.
- Biphasic (dual-wave) boluses are exported as two separate rows rather than one
  combined event. `aidR` keeps them separate but flags them, so they can be
  recombined if needed.
- CGM values are converted to mg/dL automatically where reported in mmol/L.
