---
title: "aidR: Glooko export format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{aidR: Glooko export format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(aidR)
```

## Overview

A Glooko export is a folder (possibly zipped) containing CSV files for various data types. 
A single data type may be split across two or
more files, which are then numbered `cgm_data_1.csv`, `cgm_data_2.csv`, and so
on; `aidR` reads and concatenates all matching files. Every file starts with a
one-line header such as `Name:First Last,Date Range:2024-01-01 - 2024-03-01`
before the actual column headers. Both the one-line header and the column
headers are language-specific and may differ depending on the user's account
language.

```{r}
path <- system.file("extdata", "glooko_example.zip", package = "aidR")
```

## How aidR detects this format

Because headers in Glooko exports are language-specific,
`aidR` cannot rely on header names to identify a file.
Instead, it checks the *structure* of the first line
(`something:something,something:date - date`) and distinguishes between data
types by matching the expected number of columns, which columns are numeric,
and (where available) a language-independent keyword such as `"(g)"` for carbs
or `"Insulin"` for bolus/basal/total-insulin files.

## Files and the data types they produce

```{r}
data <- parse_data(id = "1", paths = path)
```

`aidR` compiles the raw Glooko CSVs into a smaller set of data types. The
list below describes each raw file, shows an example and, in italics, notes which clean data
type(s) it is compiled into.

- **`alarms_data_*.csv`**: pump alarms.

```{r}
head(data$alarm)
```

- **`bg_data_*.csv`**: self-measured blood glucose (SMBG). *Compiled into `SMBG`.*

```{r}
head(data$SMBG)
```

- **`carbs_data_*.csv`** / **`cgm_carbs_data_*.csv`**: carbohydrate intake. These
  do **not** correspond to the carbohydrate intake communicated to the bolus
  calculator, and are almost always missing entirely. They are therefore
  ignored when building the clean `carbs` output, which instead comes from the
  bolus files (see below).
- **`cgm_data_*.csv`**: CGM data. *Compiled into `cgm`.*
```{r}
head(data$cgm)
```
- **`Insulin data/basal_data_*.csv`**: basal rates. *Compiled into `basal`.* For
  Omnipod users these are often very sparse or missing.
  For these users, the few entries that do appear correspond to basal rates delivered while in 
  manual mode; basal delivered in automated mode is unknown and can only be 
  roughly approximated from the daily aggregated insulin.
```{r}
head(data$basal)
```
- **`Insulin data/bolus_data_*.csv`**: boluses. *Compiled into `bolus`, and also
  into `carbs`* (bolus calculator). Also records whether
  a bolus was normal, square-wave, or dual-wave (although the duration for the latter two are unknown).
```{r}
head(data$bolus)
```
- **`Insulin data/insulin_data_1.csv`**: daily aggregated bolus, basal, and total
  insulin.
```{r}
head(data$total_insulin)
```
- **`Manual data/` files** (`exercise_data_*`, `food_data_*`,
  `manual_insulin_data_*`, `medication_data_*`, `notes_data_*`): self-reported
  entries made by the user in the Glooko app - exercise, food intake, manual
  insulin deliveries, medication, and free-text notes, respectively. These are
  often very sparse or missing entirely.

## Notes

- CGM and SMBG values are reported in mg/dL or mmol/L depending on account
  settings; `aidR` converts to mg/dL automatically.
- Serial numbers are reported for some data types (e.g. basal, bolus) but are
  not mapped to a device/pump name.
- `sensor_name` for CGM is always `NA`, since Glooko exports do not report it.