---
title: "aidR: Tandem Source export format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{aidR: Tandem Source export format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(aidR)
```

## Overview

A Tandem Source export is a folder (possibly zipped) containing one CSV per data
type. Filenames combine the participant, serial number, and export date with a
data-type suffix, e.g.

```
XXX_1111111_10Mar2026_1111-CGM.csv
XXX_1111111_10Mar2026_1111-Basal-doses.csv
XXX_1111111_10Mar2026_1111-Bolus.csv
XXX_1111111_10Mar2026_1111-BG.csv
XXX_1111111_10Mar2026_1111-Hourly-basal-delivery.csv
```

Column headers are always in English regardless of account language. Each file
records the device type (e.g. `Tandem Mobi™ System`) and serial number per row,
so the pump and sensor can be identified directly from the data.

```{r}
path <- system.file("extdata", "tandem_source_example.zip", package = "aidR")
```

## How aidR detects this format

Because the headers are stable and language-independent, `aidR` identifies each
file by matching its column names against a lookup of the expected headers per
data type. Note that the filename suffix is *not* used for this: a file is
recognized purely from its header row, so renamed or re-exported files are still
dispatched correctly.

## Files and the data types they produce

```{r}
data <- parse_data(id = "1", paths = path)
```

`aidR` compiles the raw Tandem Source CSVs into a smaller set of data types. The
list below describes each raw file, shows an example and, in italics, notes
which clean data type(s) it is compiled into.

- **`*-CGM.csv`**: CGM data. *Compiled into `cgm`.*
```{r}
head(data$cgm)
```
- **`*-Basal-doses.csv`**: the basal insulin dose commanded at each event.
  *Compiled into `basal`.* Tandem reports doses rather than rates and gives no
  explicit duration, so `aidR` derives the duration of each entry from the gap to
  the following one and converts the dose into a rate in U/h.
```{r}
head(data$basal)
```
- **`*-Bolus.csv`**: insulin boluses, including the bolus type and delivery
  method, the food and correction components, and the completion status.
  *Compiled into `bolus`, and also into `carbs`*, since the carbohydrates
  communicated to the bolus calculator are recorded on the same rows.
```{r}
head(data$bolus)
```
```{r}
head(data$carbs)
```
- **`*-BG.csv`**: self-measured blood glucose. *Compiled into `SMBG`.*
```{r}
head(data$SMBG)
```
- **`*-Hourly-basal-delivery.csv`**: basal insulin aggregated per hour, with one
  column per hour of the day. This is preserved as `hourly_basal` rather than
  merged into `basal`, since its wide, aggregated layout does not fit the
  per-event schema used across platforms.
```{r}
head(data$hourly_basal)
```

## Notes

- `timezone_offset` is always `NA`, since Tandem Source exports report only local
  (wall-clock) time.
- The device type is carried through as `pump_name` for insulin data and
  `sensor_name` for glucose data. Note that these strings may contain invisible
  Unicode direction marks around the device name, as they appear in the export.
- Because basal durations are inferred from consecutive timestamps, the final
  basal entry of an export has no successor and its duration is therefore
  unknown.
