---
title: "aidR: CareLink export format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{aidR: CareLink export format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(aidR)
```

## Overview

A CareLink (Medtronic) export is a *single* semicolon-delimited `.csv` (or
`.xlsx`) file. The file is divided into **sections** separated by marker rows of
the form `-------;MiniMed 780G MMT-1886;Pump;XXXXXXXXXX;-------`. Each section
has its own header row and its own set of columns:

```{r, include=FALSE}
path <- system.file("extdata", "carelink_example.csv", package = "aidR")
```

```{r, comment = ""}
writeLines(readLines(path, n = 10))
```

## How aidR detects this format

`aidR` identifies a CareLink export by the keyword `MiniMed`: CareLink only
accepts uploads from Medtronic MiniMed pumps, and MiniMed pumps can only upload
to CareLink, so the presence of `MiniMed` uniquely marks the format.

## Sections and the data types they produce

```{r}
data <- parse_data(id = "1", paths = path, clean_files = FALSE)
```

Because all information is present in one file, most of the parsing work happens up
front: the file is split into sections, and the large "Pump" section is then
split further into individual event types based on which columns are populated
on each row. (Here `clean_files = FALSE` returns these intermediate sections so
they can be shown individually.)

The list below describes each section, shows an example and, in italics, notes
which clean data type(s) it is compiled into. Sections and event types without
such a note are parsed and returned, but are not standardized further.

- **Header block** (first rows, before the first `-------`): patient, pump, and
  sensor information. `aidR` reads the pump name (e.g. `MiniMed 780G MMT-1886`)
  and sensor name (e.g. `Guardian™ 4 Sensor`) from here.

- **Pump section**: split by event type. The following are compiled into clean
  data types:

  - basal rates. *Compiled into `basal`.*

```{r}
head(data$basal)
```

  - boluses. *Compiled into `bolus`.* Records whether a bolus was normal,
    square-wave, or dual-wave, along with the delivered volume and, for extended
    boluses, the duration.

```{r}
head(data$bolus)
```

  - bolus wizard (BWZ) entries, i.e. the carbohydrates communicated to the bolus
    calculator together with the resulting recommendation. *Compiled into
    `carbs`.* `aidR` matches these against the bolus events so that a
    carbohydrate entry can be linked to the bolus it triggered.

```{r}
head(data$bwz)
```

  - self-measured blood glucose (SMBG) entries. *Compiled into `SMBG`.*

```{r}
head(data$SMBG)
```

  The remaining event types are extracted and returned, but are not compiled
  into a clean data type:

  - alerts and alarms raised by the pump.

```{r}
head(data$alerts)
```

  - prime events (tubing/cannula fills).

```{r}
head(data$prime)
```

  - temp basal events.

```{r}
head(data$temp_basal)
```

  - correction-bolus information.

```{r}
head(data$corr_bolus_info)
```

  - sensor calibrations.

```{r}
head(data$sensor_calibration)
```

  - insulin action curve changes.

```{r}
head(data$insulin_action_curve)
```

  - device change events.

```{r}
head(data$device_changes)
```

- **Aggregated Auto Insulin Data section**: daily aggregated insulin delivered
  in automated mode.

```{r}
head(data$aggr_daily_insulin)
```

- **Sensor section**: CGM data. *Compiled into `cgm`.*

```{r}
head(data$cgm)
```

## Notes

- CGM and SMBG values are reported in mg/dL or mmol/L depending on account
  settings; `aidR` converts to mg/dL automatically.
- `timezone_offset` is always `NA`, since CareLink exports report only local
  (wall-clock) time.
- Which sections are present depends on the pump/CGM combination and on what the
  participant actually used; empty sections are dropped rather than returned as
  empty data frames.

