---
title: "Introduction to aidR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to aidR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(aidR)
library(dplyr)
```

## Overview

Automated insulin delivery (AID) systems (hybrid closed-loop insulin pumps)
report their data through a variety of vendor and third-party platforms.
Every platform exports data in its own file format, with its own column
names, units, and quirks. **aidR** takes these heterogeneous exports and
turns them into a small set of standardized, tidy data frames -- one each for
continuous glucose monitoring (CGM), basal insulin, bolus insulin, and
carbohydrate intake -- so that downstream analysis code can be written once,
regardless of which platform a participant's data came from.

The typical workflow has four steps:

1. **Read and clean** every export file belonging to one participant
   (`parse_data()`, which internally calls the `clean()` generic).
2. **Merge per participant**, combining all files belonging to one person
   into one data frame per data type, and removing duplicate or invalid
   entries (`merge_per_id()`, `remove_duplicates()`).
3. **Check completeness** of the merged data, e.g. to flag missing data
   types or missing days before an analysis (`check_completeness()`,
   `check_completeness_range()`).
4. **Merge across participants** into one long-format data set for the whole
   study population (`merge_all()`).

The rest of this vignette walks through each step, describes the supported
export formats, and shows how to filter the resulting data.

## Supported export formats

`parse_data()` figures out, from the *content* of each file, which platform
it came from -- not from its file name. This means you can drop files from
several participants and several platforms into one folder (or a `.zip`
archive, including nested zips) and `parse_data()` will dispatch every file
to the right reader automatically. Files with irrelevant extensions (`.png`,
`.pdf`, `.jpg`, `.jpeg`, `.txt`, `.json`, `.bib`) are silently ignored, as are
macOS metadata files such as `.DS_Store` and `__MACOSX`.

aidR currently supports five export formats:

| Platform | Typical file type | Detected by | `clean()` method |
|---|---|---|---|
| [Glooko](https://www.glooko.com) | one CSV per data type | first line matching `Name:...,Date Range:<date> - <date>` | `clean.glooko()` |
| CareLink (Medtronic) | single `.xlsx`/`.csv` export | sheet/column layout of a CareLink export | `clean.carelink()` |
| mylife (Ypsomed) | single semicolon-delimited CSV | lines matching a `;dd.mm.yy;` timestamp pattern | `clean.mylife()` |
| [Tidepool](https://www.tidepool.org) | single multi-sheet `.xlsx` | presence of a `CGM` sheet | `clean.tidepool()` |
| YourLoops (Diabeloop) | one CSV per data type | exact column header match against a known lookup table | `clean.yourloops()` |

Each reader (`read_glooko()`, `read_carelink()`, `read_mylife()`,
`read_tidepool()`, `read_yourloops()`) returns the data types found in that
particular file, tagged with an S3 class matching the platform. `clean()`
is an S3 generic dispatching on that class, and standardizes each recognized
data type to a common column layout:

- **`cgm`**: `timestamp`, `value` (mg/dL)
- **`basal`**: `timestamp`, `duration`, `amount`, `rate`
- **`bolus`**: `timestamp`, `type`, `sub_type`, `duration`, `extended`, `normal`
- **`carbs`**: `timestamp`, `carbs_grams`, `label`, `estimated_absorption_duration`

Platforms often report additional data beyond these four core types (e.g.
CareLink also exposes sensor calibrations, alerts, temp basal and device
change events; Tidepool exposes pump settings sheets; Glooko exposes total
daily insulin, SMBG and alarms). These extra types are preserved in the
output under their own name, just not standardized across platforms, since
their availability and structure vary by platform.

Not every platform reports every core type: for example, mylife exports only
contain `bolus` and `carbs` entries.

## Reading and cleaning data for one participant

`parse_data()` takes a participant ID and one or more paths (files, folders,
or zip archives) and returns a named list, one entry per file found, with
the cleaned data for that file:

```{r}
path <- system.file("extdata", "1_cycle_1_arm_1_ms_aid_data_1.xlsx", package = "aidR")

data_101 <- parse_data(id = "101", paths = path)
str(data_101, max.level = 1)
str(data_101[[1]], max.level = 1)
```

Here, `path` happens to be a single Tidepool export, so `data_101` has one
entry, containing the standardized `cgm`, `basal`, `bolus` and `carbs` data
frames alongside Tidepool-specific extras (`device_event`, `food`,
`basal_schedules`, ...).

Passing `clean_files = FALSE` returns the raw, unstandardized data as read
from the file, tagged with its platform class -- useful for debugging a
parser or inspecting a platform's raw layout.

## Merging data per participant

A participant's data commonly spans several files (e.g. repeated exports
over time, or exports from more than one platform). `merge_per_id()` binds
all files belonging to one participant together, by data type, and sorts
each resulting data frame by `timestamp`:

```{r}
merged_101 <- merge_per_id(data_101)
str(merged_101, max.level = 1)
head(merged_101$cgm)
```

Raw merged data commonly still contains duplicate or implausible entries --
for instance, overlapping exports reporting the same CGM reading twice, or
sensor noise outside the physiologically plausible range.
`remove_duplicates()` removes exact duplicate rows from every data type, and
additionally applies CGM-specific cleanup: truncating implausible values and
downsampling to a regular time interval (5 minutes by default) when the same
period is covered by more than one reading:

```{r}
nrow(merged_101$cgm)

clean_101 <- remove_duplicates(merged_101)
nrow(clean_101$cgm)
```

All three thresholds can be adjusted or disabled (`truncate_min_cgm`,
`truncate_max_cgm`, `min_minutes_difference_cgm`; set to `NULL` to skip that
particular step).

## Checking data completeness

Before running an analysis, it is often useful to verify that a participant
actually has data of every type you need, and that it is not missing entire
days. `check_completeness()` warns if a data type is absent, or present but
empty:

```{r}
check_completeness("101", clean_101)
```

`check_completeness_range()` goes a step further and checks, for a given
date range, which specific days have no data at all for each type -- handy
for spotting gaps such as a sensor or pump disconnecting for a few days:

```{r}
date_range <- range(as.Date(clean_101$cgm$timestamp))
date_range

check_completeness_range(
  "101", clean_101,
  first_date = as.character(date_range[1]),
  last_date  = as.character(date_range[2])
)
```

Both functions accept a `necessary_types` argument to restrict (or extend)
which data types are checked, and `allow_total_insulin_instead_of_basal`,
which avoids spurious warnings for pumps (e.g. Omnipod) that only report
`total_insulin` rather than `basal` rates while in manual mode.

## Merging data across participants

Once every participant's data has been parsed, merged and cleaned,
`merge_all()` combines them into one long-format data set per data type,
adding an `id` column so participants can be told apart:

```{r}
# For illustration, we reuse the same sample export for a second participant
data_102  <- parse_data(id = "102", paths = path)
clean_102 <- remove_duplicates(merge_per_id(data_102))

all_data <- list("101" = clean_101, "102" = clean_102)
combined <- merge_all(all_data)

str(combined, max.level = 1)
head(combined$cgm)
```

`merge_all()` always returns the four core types (`cgm`, `basal`, `bolus`,
`carbs`); a participant missing one of them simply contributes no rows to
that type.

## Filtering data

Because every standardized data type is a plain tibble with a `timestamp`
column, filtering uses ordinary `dplyr`/base R tools -- no special aidR
function is required. A few common cases:

**Filter to one participant** (after `merge_all()`):

```{r}
combined$cgm %>%
  filter(id == "101") %>%
  head()
```

**Filter to a date range**, e.g. the first three days of CGM data:

```{r}
combined$cgm %>%
  filter(timestamp >= min(timestamp), timestamp < min(timestamp) + lubridate::ddays(3)) %>%
  nrow()
```

**Filter by a value condition**, e.g. hypoglycemic CGM readings (< 70 mg/dL):

```{r}
combined$cgm %>%
  filter(value < 70) %>%
  head()
```

**Filter by category**, e.g. only automated (closed-loop) boluses:

```{r}
combined$bolus %>%
  filter(type == "automated") %>%
  head()
```

CGM-specific filtering that is common enough to warrant a dedicated
implementation (truncating out-of-range values and downsampling to a
regular interval) is already built into `remove_duplicates()`, described
above.

## Putting it all together

A typical study-level pipeline combines all of the above for every
participant in a loop:

```{r, eval = FALSE}
ids   <- c("101", "102")           # participant IDs
paths <- list(                      # one or more export paths per participant
  "101" = "path/to/participant_101_export.zip",
  "102" = "path/to/participant_102_export.zip"
)

all_data <- list()
for (id in ids) {
  parsed          <- parse_data(id, paths[[id]])
  merged          <- merge_per_id(parsed)
  all_data[[id]]  <- remove_duplicates(merged)

  check_completeness(id, all_data[[id]])
}

combined <- merge_all(all_data)
```

`combined` can then be filtered, summarized, or exported for statistical
analysis exactly as shown in the previous section.
